# Multiple Linear Regression, Part 4 in SAS

::: {.graybox}

<font size="4">**Goals**</font>

The goal of this software lesson is to learn how to

* create graphical summaries (i.e., interaction plot) when there is a potential interaction, and 
* include interaction terms in a model and obtain model-based estimates.

:::

## Dataset  

Let’s use the Lung Health Study (LHS) again. Recall this was the dataset used in software lessons in the previous course (PubH 6450: Biostatistics 1).

Recall that the dataset and data dictionary can be found on Canvas, *lhs.csv* and *lhsdoc.docx*, respectively.

```{r, eval=FALSE, echo=TRUE, comment = ""}
proc import
  DATAFILE = "/<filepath>/lhs.csv"
  OUT = lhs
  DBMS = csv
  REPLACE;
  GETNAMES = yes;
run;
```

## Interactions

Thus far, you have only dealt with models that are additive (e.g., $y = b_0 + b_1x_1 + b_2x_2$), with continuous or categorical predictors. If there is a binary predictor and a continuous predictor in an additive model, the two groups have similar slopes but different intercepts. However, it may be the case that the two groups have different intercepts AND different slopes. This is a situation of an interaction between the predictors (e.g., $y = b_0 + b_1x_1 + b_2x_2 + b_3x_1x_2$).  

In general, interactions can be between any type of predictors: continuous, binary, or multicategorical. This software lesson will present the situation of an interaction term between one binary and one continuous predictor, as well as two binary predictors. The former situation will be presented first followed by the latter.

### Interaction between one binary and one continuous predictor

Suppose the Lung Health Study researchers were interested in the following research question: 

<center>
*Does the effect of intervention (smoking intervention vs. usual care) on lung function at 5th year depend on cigarettes smoked per day at baseline?*
</center><br>

When dealing with a potential interaction situation, it's helpful to visualize the data prior to obtaining the regression results from the interaction model. 

However, before you proceed, notice that the question is about comparing two groups, and not three (as in the original intervention variable). Therefore, to answer this question, you first have to create a new variable for the intervention variable that takes the three interventions (SI-A, SI-P, UC) and collapses the smoking interventions into a single group (SI) so that you are left with two interventions (SI, UC). 

### Collapse categories in a categorical variable

- To collapse categories of a categorical predictor, 
    + use the `data` step, specifying the name of the new data set in the first line of code (e.g., `lhs1`), then specifying the name of the data set that you want to make a copy of in the `SET` statement (e.g., `lhs`). 
    + Then use the `IF...THEN` statement and `ELSE` statement, telling SAS how to create the categories in the new variable (`groups2`) from an existing numerical variable in the data set (`alphagroup`):

```{r, eval=FALSE, echo=TRUE, comment = ""}
data lhs1;
SET lhs;
IF alphagroup = "SI-A" THEN groups2 = "SI";
ELSE IF alphagroup = "SI-P" THEN groups2 = "SI";
ELSE groups2 = "UC";
run;
```

Be sure to check that the code carried out the task as you intended (e.g., `proc freq` with the `TABLE` statement of the two variables, `alphagroup` and `groups2`, to check that the counts of SI in the new variable equals the sum of SI-A and SI-P from the original variable).

### Create interaction plot & Fit a model with interaction term

For situations that involve a two-variable interaction, you can visualize the interaction by creating an interaction plot. There are two different ways of plotting this, depending on the types of variables that are interacting.

In SAS, when you fit an interaction model with a categorical predictor and a continuous predictor, in addition to the model results (e.g., estimate, SE, CI, and *p*-value), `proc glm` also produces an interaction plot (two for the price of one!). 

- To create an interaction plot for a categorical predictor and a continuous predictor and fit an interaction model, 
    + `proc glm`, specifying the name of the dataset and (optional) `PLOTS(MAXPOINTS=NONE)` to produce an interaction plot when sample size is larger than 5,000 (Note: This `PLOTS` option is only needed for large sample sizes, otherwise SAS does it automatically),
    + the `CLASS` statement with all the categorical variables in the model that are not 0/1 and setting the reference level using `(ref="value")` (if applicable), and
    + the `MODEL` statement with the form `y = x1 x2 x1*x2` that includes all of the variables for the model but putting a star, `*`, between the variables that are interacting. Also include the options of `solution` to get obtain the table with the regression coefficient estimates and inferential results, `clparm` to obtain confidence intervals for the regression coefficients, and `alpha=` to specify the confidence level: 

```{r, eval=FALSE, echo=TRUE, comment = ""}
proc glm data = lhs1 PLOTS(MAXPOINTS=NONE);
CLASS groups2(ref="UC");
MODEL FEVFVC5 = f10cigs groups2 f10cigs*groups2 / solution clparm alpha = 0.05;
run;
```

The interaction plot is at the bottom of the output. Estimated regression lines are displayed for each group. 

Additionally, the regression tables are similar to the output shown previously, except now you have a row for the interaction term between `groups2` and `f10cigs`. You should examine the significance of the interaction term to determine what the next step will be.

**Interaction Significant? Obtain slopes of the continuous predictor for each category**

If an interaction exists, you can obtain the continuous predictor's slope for each category. 

Although the interaction between intervention group and number of cigarettes in this scenario isn't significant, the code for obtaining this information will be presented for demonstration purposes. 

To estimate these group effects, you use the `ESTIMATE` statement in `proc glm`, specifying the terms that contribute to calculating the slope (i.e., the continuous predictor term and the interaction term). For the continuous predictor, you list the name of that variable and then put a `1` after it. Then for the interaction term, you list the interaction term from the model and then indicate `1 0` (if you want the slope for the group of interest) or `0 1` (if you want the slope for the reference group). Then, the software will compute the slopes for each group for you, and it's associated statistics (e.g., SE, CI, *p*-value). While you could compute the slopes by hand for each group, the other statistics would be harder to compute by hand.

- To obtain group slopes when an interaction is significant, use the code you have seen thus far (the `proc glm` task) but also use two `ESTIMATE` statements, one for each slope, and including: 
    + `"<insert label for first group slope>"` to specify the label in the table, 
    + `<cont var> 1` to specify the name of the continuous variable (e.g., `f10cigs 1`), and 
    + `<interaction term> 1 0` to compute a slope for the group of interest or `<interaction term> 0 1` to compute a slope for the reference group:

```{r, eval=FALSE, echo=TRUE, comment = ""}
proc glm data = lhs1;
CLASS groups2(ref="UC");
MODEL FEVFVC5 = f10cigs groups2 f10cigs*groups2 / solution clparm alpha = 0.05;
ESTIMATE "Slope for f10cigs in SI" f10cigs 1 f10cigs*groups2 1 0;
ESTIMATE "Slope for f10cigs in UC" f10cigs 1 f10cigs*groups2 0 1;
run;
```

In the second to last table of the output, the relevant information for group slopes are their estimates (`Estimate`), SEs (`Standard Error`), *p*-values (`Pr > |t|`), and CIs (`95% Confidence Limits`).

**Interaction NOT Significant? Obtain adjusted slopes for each main effect**

Recall that if the interaction term is not statistically significant, you might consider dropping it from the model and fit a main effects model instead. Note: This week is the first time you've seen the phrase "main effects", which essentially means predictors that are not interacting with one another and their regression coefficients are thus adjusted for the other variables in the model. When a model doesn't have an interaction term in it, it's called a "main effects model". 

- To specify a main effects model, use the code you have seen thus far, but drop the interaction term in the model:

```{r, eval=FALSE, echo=TRUE, comment = ""}
proc glm data = lhs1;
CLASS groups2(ref="UC");
MODEL FEVFVC5 = f10cigs groups2 / solution clparm alpha = 0.05;
run;
```

The output is similar to the output shown previously.

### Interaction between two binary predictors

Unlike in the previous interaction scenario, where there was a slope for a continuous variable for each group, when there is an interaction term with two binary predictors, you obtain group means and can examine the pairwise difference between group means.

Now suppose the Lung Health Study researchers were interested in the following research question: 

<center>
*Does the effect of intervention (smoking intervention vs. usual care) on lung function at 5th year depend on biological sex?*
</center><br>

Let's go through a similar approach as show in the previous scenario. 

### Create interaction plots 

- To create an interaction plot for two interacting categorical predictors, 
    + `ODS SELECT InteractionPlot` to restrict the output to just the interaction plot,
    + `proc genmod`, specifying the name of the dataset, 
    + the `CLASS` statement with all the categorical variables in the model and setting the reference level using `(ref="value")` (if applicable), 
    + the `MODEL` statement with the form `y = x1 x2 x1*x2` that includes all of the variables for the model but putting a star, `*`, between the variables that are interacting, and 
    + the `EFFECTPLOT` statement with `interaction` to create an interaction plot:

```{r, eval=FALSE, echo=TRUE, comment = ""}
ODS SELECT InteractionPlot;
proc genmod data=lhs1;
CLASS groups2(ref="UC") AGENDER;
MODEL FEVFVC5 = groups2 AGENDER groups2*AGENDER;
EFFECTPLOT interaction;
run;
```

Lines and points (which are the observed group means) are displayed for each group combination. 

### Fit a model with interaction term

You already saw how to fit a model with the interaction in the previous scenario. Let's use the same code, but substitute in the variables of interest to answer this particular research question. 

```{r, eval=FALSE, echo=TRUE, comment = ""}
proc glm data = lhs1;
CLASS groups2(ref="UC") AGENDER;
MODEL FEVFVC5 = groups2 AGENDER groups2*AGENDER / solution clparm alpha = 0.05;
run;
```

The output is similar to the output shown previously, except now you have a row for the interaction term between `groups2` and `AGENDER`. You should examine the significance of the interaction term to determine what the next step will be.

**Interaction Significant? Obtain group means and pairwise group mean differences** 

If an interaction exists, you can obtain statistics for the group means and also differences between the group means. 

Although the interaction between intervention group and sex in this scenario aren't significant, the code for obtaining this information will be presented for demonstration purposes. 

- To obtain group means and pairwise differences between group means when an interaction is significant, use the code you have seen thus far (the `proc glm` task) but also use the `LSMEANS` statement with the two interacting predictors, separated by a star, `*`, and including the options of
    + `cl` to compute confidence intervals for the group means and pairwise differences between the group means, 
    + `adjust=tukey` to use adjust the inferential results for multiple comparisons using Tukey method,
    + `pdiff` to get the *p*-values for testing the pairwise differences, and 
    + `stderr` to compute the standard errors for the group means:

```{r, eval=FALSE, echo=TRUE, comment = ""}
proc glm data = lhs1 PLOTS(MAXPOINTS=NONE);
CLASS groups2(ref="UC") AGENDER;
MODEL FEVFVC5 = groups2 AGENDER groups2*AGENDER / solution clparm alpha = 0.05;
LSMEANS groups2*AGENDER / cl adjust=tukey pdiff stderr;
run;
```

The relevant output for group means and pairwise difference in group means are all the way at the bottom of the output, under the section that says `Least Squares Means`. 

The relevant information in the first table in this section is the group mean (`LSMEAN`) and the model-estimated standard errors (`Standard Error`). It also gives each group combination a number (`LSMEAN Number`) that is to be referenced in the subsequent tables.

The second table displays the *p*-values for testing differences between the group means. 

The third table has the lower and upper values for the 95% confidence intervals for the group means (`95% Confidence Limits`). 

And the fourth table presents the `Difference Between Means` value as well as the lower and upper values for the 95% confidence interval (`Simultaneous 95% Confidence Limits for LSMean(i)-LSMean(j)`) for the differences.

**Interaction NOT Significant? Obtain adjusted means and average differences between adjusted means**

Recall that if the interaction term is not statistically significant, you might consider dropping it from the model and fit a main effects model instead. 

- To specify a main effects model, use the code you have seen thus far, but drop the interaction term in the model:

```{r, eval=FALSE, echo=TRUE, comment = ""}
proc glm data = lhs1;
CLASS groups2(ref="UC") AGENDER;
MODEL FEVFVC5 = groups2 AGENDER / solution clparm alpha = 0.05;
LSMEANS groups2 AGENDER / cl adjust=tukey pdiff stderr;
run;
```

The relevant output for adjusted means and average difference in adjusted means (i.e., main effect for each predictor) are all the way at the bottom of the output again, under the section that says `Least Squares Means`. 

The output is similar to before, except that the results are presented for each predictor individually. Also, the first table for the predictors will include an extra two columns (under the column spanner `H0:LSMean1=LSMean2`) for testing the differences between the groups in that variable. 