# Logistic Regression in SAS

::: {.graybox}

<font size="4">**Goals**</font>

The goal of this software lesson is to learn how to

* fit a logistic regression model and obtain estimates for the effect of a predictor in terms of log odds and in terms of odds ratios, and
* obtain model diagnostics to check assumptions and evaluate the model fit.

:::

## Dataset 

In this software lesson, you will use the Core Temperature dataset. The dataset, data dictionary, and dataset introduction can be found on Canvas, titled *Core-Temp.csv*, *Core-Temp-Data-Dictionary.pdf*, and *Core-Temp-Dataset-Intro.pdf*, respectively.

```{r, eval=FALSE, echo=TRUE, comment = ""}
proc import
  DATAFILE = "/<filepath>/Core-Temp.csv"
  OUT = core_temp
  DBMS = csv
  REPLACE;
  GETNAMES = yes;
run;
```

## Logistic Regression

Thus far in the course, you have focused on *linear* regression, which is used when you have a continuous outcome. But, if the outcome is not continuous, other types of regression methods are necessary. This software lesson builds on what you have learned thus far, but focuses on *logistic* regression, which is used when you have a binary outcome. 

In the Core Temperature study, the researchers were interested in examining the relationship between the development of serious infections (`SeriousInfection`) and time weighted average core temperature (`TWATemp`) in Celsius, after adjusting for `Age` at time of surgery (in years), sex (`FEMALE`), body mass index at time of surgery (`BMI`), and Charlson score (`CharlsonScore`). In this example, `TWATemp` is the predictor of interest, and `Age`, sex, `BMI`, and `CharlsonScore` are considered to potential confounders.

### Fit logistic regression model

When you fit a logistic regression model, you can obtain regression coefficient results in terms of log odds or odds ratios. The model is fit on the log-odds scale, but the results are often more interpretable when the values are converted to odds ratios.

*Estimates in terms of log odds and odds ratios*

- To fit a logistic regression model and obtain log odd estimates, use
    + `proc logistic`, specifying the name of the dataset, 
    + the `CLASS` statement with all the categorical variables in the model and setting the reference level using `(ref="value")` (if applicable), and
    + the `MODEL` statement with the form `y(event="value") = x1 x2 x3`, denoting the event of interest in quotation marks for the response variable and separating the x-variables with a space and including the options of `cl` to get obtain confidence intervals: 
    
```{r, eval=FALSE, echo=TRUE, comment = ""}
proc logistic data = core_temp;
CLASS FEMALE(ref="0") / param=ref; 
MODEL SeriousInfection(event="1") = TWATemp Age FEMALE BMI CharlsonScore / cl; 
run;
```

Note: Predictor variables that are already coded as indicator variables (such as FEMALE, which is coded 1 if female and 0 if not) do not need to be included in the CLASS statement at all. But if they are included, then be sure to set ref="0"; otherwise SAS by default will recode them with the last level (1 in this case) as the reference level. 

The output provides twelve tables. The tables of interest for this week are in ninth (titled `Analysis of Maximum Likelihood Estimates`), tenth (titled `Odds Ratio Estimates`), and twelfth (titled `Parameter Estimates and Wald Confidence Intervals`) tables.

The `Analysis of Maximum Likelihood Estimates` (ninth) table provides the information for the regression coefficients, in terms of log-odds, including:

 - the regression coefficient values in terms of log odds under the column `Estimate`, with the intercept denoted in the first row and the regression coefficients denoted in the subsequent rows. For example, the regression coefficient value for `TWATemp` is 0.230, which means that for every one-degree (in Celsius) increase in core temperature, the log odds of a developing serious infection increases by 0.230 (holding all else constant).

- the *p*-values for testing the regression coefficients in the column `Pr > ChiSq`. 

The `Odds Ratio Estimates` (tenth) table displays the odds ratios (`Point Estimate`) and their confidence intervals (`95% Wald Confidence Limits`) for each of the predictors. For example, the regression coefficient value for `TWATemp` is 1.2587 (which is $exp(0.230)$), which means that for every one-degree (in Celsius) increase in core temperature, the odds of a developing serious infection increases by 25.87% (holding all else constant).

The `Parameter Estimates and Wald Confidence Intervals` (twelfth) table presents the log-odds estimates (again) (`Estimate`) and their confidence intervals (`95% Confidence Limits`) for each of the predictors.

::: {.bluebox}

<font size="4">**Additional Resources**</font>

For additional SAS Resources on logistic regression, see:

* <a href="https://stats.idre.ucla.edu/sas/dae/logit-regression/"> Logit Regression </a> (UCLA)
* <a href="https://stats.idre.ucla.edu/sas/output/proc-logistic/"> PROC LOGISTIC SAS Annotated Output </a> (UCLA)
* <a href="https://support.sas.com/documentation/cdl/en/statug/63033/HTML/default/viewer.htm#statug_logistic_sect049.htm"> ODS Table Names in PROC LOGISTIC </a> (SAS)
* <a href="https://www.youtube.com/watch?v=DKWoM_Greyk"> Saving Predicted Probability in PROC Logistic VIDEO </a> (Analytics University)
* <a href="https://support.sas.com/documentation/cdl/en/statug/63962/HTML/default/viewer.htm#statug_logistic_sect018.htm"> Output statement in PROC LOGISTIC </a> (SAS)

:::

## Model Evaluation

As with any statistical method, you need to check model assumptions and evaluate model fit. Some code from previous software lessons also apply for logistic regression: 

- `proc corr` for checking multicollinearity (Note: You can't get VIF values using `proc logistic` in SAS)

Only new model diagnostic topics are presented in this software lesson.

*Linearity of continuous predictors*

One of the assumptions for logistic regression is **L**inearity, meaning the continuous predictor(s) needs to be linearly related to the *log odds* of the binary outcome. You can check this by examining a scatterplot of the model-predicted log odds (i.e., logit) verses the continuous predictor and look to see if the form is approximately linear. 

- To create scatterplot(s) of model-predicted logit vs. continuous predictor(s), 
    + first, use the code you have seen thus far, include the option `NOPRINT` to suppress displaying the results and use the `OUTPUT` statement, specifying the name of the new dataset in `out=` , and the statistics you want to save (e.g., `predicted` for probabilities) followed by how you want to name of the variable after the equals sign. 
    + Then, create a copy of the `model1` dataset and create a new variable that computes the log-odds (i.e., logit) for each observation (`logit = log(probabilities/(1-probabilities))`).
    + Then, for each predictor, use `proc sgplot`, specifying the name of the dataset, followed by the `LOESS` statement, specifying a continuous predictor for the x-axis (e.g., `x = TWATemp`), the logits for the y-axis (`y = logit`), and the option to modify the line color (`lineattrs = (color = "blue")`), and finally the `REG` statement with similar structure to `LOESS` (but changing the line color to red):
        
```{r, eval=FALSE, echo=TRUE, comment = ""}
*Saving the model-predicted probabilities to a new dataset; 
proc logistic data = core_temp NOPRINT;
CLASS FEMALE(ref="0") / param=ref; 
MODEL SeriousInfection(event="1") = TWATemp Age FEMALE BMI CharlsonScore / cl; 
OUTPUT out = model1 predicted = probabilities;
run;

*Calculating the model-predicted log-odds (logit) values; 
data model1;
  SET model1;
  logit = log(probabilities/(1-probabilities));
run;

*Scatterplot of logit vs. TWATemp; 
ods GRAPHICS / LOESSOBSMAX = 8000; *Needed only for datasets larger than 5000; 
proc sgplot data = model1;
LOESS x = TWATemp y = logit / lineattrs = (color = "blue");
REG x = TWATemp y = logit / lineattrs = (color = "red");
run;

*Scatterplot of logit vs. Age; 
ods GRAPHICS / LOESSOBSMAX = 8000; *Needed only for datasets larger than 5000;
proc sgplot data = model1;
LOESS x = Age y = logit / lineattrs = (color = "blue");
REG x = Age y = logit / lineattrs = (color = "red");
run;

*Scatterplot of logit vs. BMI; 
ods GRAPHICS / LOESSOBSMAX = 8000; *Needed only for datasets larger than 5000;
proc sgplot data = model1;
LOESS x = BMI y = logit / lineattrs = (color = "blue");
REG x = BMI y = logit / lineattrs = (color = "red");
run;

*Scatterplot of logit vs. CharlsonScore; 
ods GRAPHICS / LOESSOBSMAX = 8000; *Needed only for datasets larger than 5000;
proc sgplot data = model1;
LOESS x = CharlsonScore y = logit / lineattrs = (color = "blue");
REG x = CharlsonScore y = logit / lineattrs = (color = "red");
run;
```

The output is scatterplots of logit vs. continuous predictor(s), one scatterplot for every continuous predictor in the model. For the Core Temperature study example, there were four continuous predictors in the model, so four scatterplots are created. 

The blue curve is a "best-fit" curve to the data, called loess - a smoothing curve - and doesn't assume any sort of form/model of the data. The red line is a "best-fit" *linear* line to the data, where the model is assumed to be a straight line. To evaluate the the linearity of the continuous predictors, you compare the blue curve to the red line in each of the plots. If the two follow a similar trajectory or are relatively the same, then the linearity assumption is met. If the blue curve has a different form (e.g., curved) than red line, then the linearity assumption isn't met for that predictor.

*Model Fit: Likelihood ratio test and AIC*

You can use likelihood ratio test (LRT) and AIC to compare model fit between several models. `Proc logistic` automatically provides you with the LRT result as well as the AIC for the model.

The AIC value is found in the sixth table (titled `Model Fit Statistics`). SAS presents the AIC for two models, one for `Intercept Only` and one for the "full" model (the one you specified) in the `Intercept and Covariates` column. (Recall that this statistic is best used when you are comparing models to see which model fits the data better. And in this case, you can compare the specified model with a model with no predictors ("intercept-only")).

The LRT is found in the seventh table (titled `Testing Global Null Hypothesis: BETA=0`), in the `Liklihood Ratio` row. It is an omnibus test, comparing the specified model to a model with only the intercept value. (Note: LRT can also be used more generally to compare different subsets of models. This topic will be discussed later in this course.)

*Hosmer-Lemeshow Goodness-Of-Fit test*

One test that can be used to check for the goodness-of-fit of the logistic regression model is the Hosmer-Lemeshow test. The Hosmer-Lemeshow test provides an overall or global assessment of how well the (logistic) model fits the data. The null hypothesis is that the model fits the data well. 

- To conduct the Hosmer-Lemeshow test, use the code you have seen thus far, but add the `lackfit` option in the `MODEL` statement:

```{r, eval=FALSE, echo=TRUE, comment = ""}
proc logistic data = core_temp;
CLASS FEMALE(ref="0") / param=ref; 
MODEL SeriousInfection(event="1") = TWATemp Age FEMALE BMI CharlsonScore / cl lackfit; 
run;
```

Two additional tables are added to the bottom of the output. The last table (the fourteenth one, titled `Hosmer and Lemeshow Goodness-of-Fit Test`) displays the test statistic (`Chi-Square`), the degrees of freedom (`DF`), and the *p*-value for the test (`Pr > ChiSq`). 

*Influential observations: DFBETAs*

Recall that you can evaluate influential observations by examining the Cook's distance and DFBETAs for each observation. In SAS, you can create plots of the DFBETAs verses observation number for a logistic model, but unfortunately, one cannot compute Cook's distance values.

- To create a plot of DFBETAs values verses observation number, use the code you have seen thus far, but add the `PLOT(ONLY)` option to `proc logistic`, specifying `DFBETAS` for the DFBETAs plot (Note: You only need to include `MAXPOINTS=NONE` for datasets that have more than 5,000 observations):

```{r, eval=FALSE, echo=TRUE, comment = ""}
proc logistic data = core_temp PLOTS(ONLY MAXPOINTS=NONE) = DFBETAS;
CLASS FEMALE(ref="0") / param=ref; 
MODEL SeriousInfection(event="1") = TWATemp Age FEMALE BMI CharlsonScore / cl; 
run;
```

The plots are displayed at the bottom of the output, one for each regression coefficient, including the intercept. For example, for the specified model for this software lesson, there are five predictors in the model, so that means there are six plots created (five plots, one for each predictor, and one for the intercept term).

::: {.bluebox}

<font size="4">**Additional Resources**</font>

For additional SAS Resources on diagnostics for logistic regression, see:

* <a href="https://www.lexjansen.com/wuss/2013/135_Paper.pdf"> A Tutorial on PROC LOGISTIC </a> (Arthur Li)

:::
